---
title: "vcfR_plot"
author: "ashar"
date: "2026-06-20"
output: html_document
---
# A. Pacakages preparation
## 1. Install packages plotVCF
```{r}
BiocManager::install("DelayedMatrixStats")
install.packages("VariantAnnotation")
install.packages("ggplot2")
install.packages("vcfR")
install.packages("withr")

```

## 2. Load vcfR library

```{r}
library(VariantAnnotation)
library(ggplot2)
library(vcfR)
library(DelayedMatrixStats)
library(withr)
library(gridExtra)
```
# B.File load

## 1. Import file for analysis (.vcf, .gff, .fasta)
```{r}
vcf <- read.vcfR("F:/materi_wgs_inbio/wgs/results/snpEff/variants_annotated.vcf")

```
## 2. Data clean
```{r}
vcf_filtered <- vcf[vcf@fix[, "CHROM"] != "NZ_OW968330.1", ] #Remove plasmid contig
```

## 3. Convert into data frame
```{r}
df <- as.data.frame(getFIX(vcf_filtered))
df$POS <- as.numeric(df$POS)
df$QUAL <- as.numeric(df$QUAL)
```

# C. Make plot
## 1. Manhattan plot
```{r}
manhattan_plot <- ggplot(df, aes(x = POS, y = QUAL, color = CHROM)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~CHROM, scales = "free_x") +
  theme_minimal() +
  labs(x = "Position", y = "Quality", title = "Variants - Mycobacterium tuberculosis")
```

## 2. Pie chart

### a. Extract ANN field from INFO field
```{r}
info      <- getINFO(vcf)
ann_raw   <- regmatches(info, regexpr("ANN=[^;]+", info))
ann_clean <- sub("ANN=", "", ann_raw)
 
# Split each variant's annotations by comma, explode ALL transcripts
all_ann   <- unlist(strsplit(ann_clean, ","))
ann_parts <- strsplit(all_ann, "\\|")
 
annotation <- sapply(ann_parts, function(x) if (length(x) >= 2) trimws(x[2]) else NA_character_)
impact     <- sapply(ann_parts, function(x) if (length(x) >= 3) trimws(x[3]) else NA_character_)

cat("\n=== Raw unique annotation values (first 20) ===\n")
print(head(sort(unique(annotation)), 20))
 
cat("\n=== Raw unique impact values ===\n")
print(sort(unique(impact)))
```


### b. Categorise SNP type
```{r}
snp_type <- dplyr::case_when(
  grepl("missense_variant",               annotation, ignore.case = TRUE) ~ "Missense",
  grepl("synonymous_variant",             annotation, ignore.case = TRUE) ~ "Silent",
  grepl("stop_gained|stop_lost|nonsense", annotation, ignore.case = TRUE) ~ "Nonsense",
  TRUE ~ "Other"
)
```

### c. Categorise impact
```{r}
impact_cat <- dplyr::case_when(
  impact == "MODIFIER" ~ "Modifier",
  impact == "LOW"      ~ "Low",
  impact == "MODERATE" ~ "Moderate",
  impact == "HIGH"     ~ "High",
  TRUE ~ NA_character_
)
```

### c. Count and calculate percentage 
```{r}
make_pct_df <- function(vec, levels_order) {
  vec <- vec[!is.na(vec)]
  tbl <- table(factor(vec, levels = levels_order))
  df  <- as.data.frame(tbl, stringsAsFactors = FALSE)
  colnames(df) <- c("label", "count")
  df$pct   <- round(df$count / sum(df$count) * 100, 2)
  df$label <- factor(df$label, levels = levels_order)
  df
}
 
snp_df    <- make_pct_df(snp_type,   c("Missense", "Silent", "Nonsense"))
impact_df <- make_pct_df(impact_cat, c("Modifier", "Low", "Moderate", "High"))
 
cat("\n=== SNP Type Counts ===\n");  print(snp_df)
cat("\n=== Impact Counts ===\n");    print(impact_df)
```
### d. Pie chart function
```{r}
make_pie <- function(df, colors, title) {
  df$legend_label <- paste0(df$label, " (", df$pct, "%)")
  ggplot(df, aes(x = "", y = count, fill = label)) +
    geom_col(width = 1,linewidth = 0.6) +
    coord_polar(theta = "y", start = 0) +
    scale_fill_manual(
      values = colors,
      labels = df$legend_label,
      breaks = as.character(df$label)
    ) +
    labs(title = title, fill = NULL) +
    theme_void() +
    theme(
      plot.title       = element_text(hjust = 0.5, size = 12,
                                      face = "bold", margin = margin(b = 8)),
      legend.position  = "right",
      legend.text      = element_text(size = 9),
      legend.key.size  = unit(0.5, "cm"),
      plot.margin      = margin(10, 10, 10, 10)
    )
}
```

### e. Color palette
```{r}
snp_colors <- c(
  "Missense" = "#4a7aad",
  "Silent"   = "#89c4e1",
  "Nonsense" = "#1a3a5c"
)
 
impact_colors <- c(
  "Modifier" = "#f2c4c8",
  "Low"      = "#d4a0a8",
  "Moderate" = "#f0dcc8",
  "High"     = "#8b1a2a"
)

```

### f. Build plot
```{r}
p_snp    <- make_pie(snp_df,    snp_colors,    "SNP Type")
p_impact <- make_pie(impact_df, impact_colors, "Variant Impact")
piechart <- arrangeGrob(p_snp, p_impact, ncol = 2)
```
## 2. Export plot


```{r}
# Manhattan plot
png("F:/materi_wgs_inbio/wgs/plot/manhattan_plot.png", 
    res = 300, width = 3000, height = 4000, units = "px", bg = "white")
print(manhattan_plot)
dev.off()
```
```{r}
# Piechart
ggsave("F:/materi_wgs_inbio/wgs/plot/pie_charts.png", piechart, width = 12, height = 8, dpi = 300, bg = "white")
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
